---
title: "Machine Learning Peer Assessment Project"
author: "Tim Gaertner"
date: "March 16, 2015"
output:
  html_document:
    fig_caption: yes
    number_sections: yes
    toc: yes
---


# Download and Read-In Raw Data
```{r, echo=F}
setwd("C:\\Users\\tgaertner\\Desktop\\Coursera\\8. Practical Machine Learning\\Week 3\\Practical-Machine-Learning")
```

* Download and Read in Training Data
```{r cache=T, message=F}
fileUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
if (!file.exists("training.csv")) {download.file(url = fileUrl, destfile = "training.csv")}
trainingData <- read.csv(file = "training.csv")
```

* Download and Read in Testing Data
```{r cache=T, message=F}
fileUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
if (!file.exists("testing.csv")) {download.file(url = fileUrl, destfile = "testing.csv")}
testingData <- read.csv(file = "testing.csv")
```

# Pre-Processing
* Clean `cvtd_timestamp` variable by formatting as a date
* Format all number based variables as numeric
* Remove variables `X`,`user_name`, and `new_window` from model consideration as they cannot be used to score the testing data

```{r}
trainingData$cvtd_timestamp <- as.Date(trainingData$cvtd_timestamp,"%m/%d/%Y %H:%M")

for (i in 1:length(names(trainingData))) {
        if (grepl("dumbbell|forearm|arm|belt",names(trainingData)[i])){
                trainingData[,i] <- as.numeric(trainingData[,i])
        }
}

trainingData <- trainingData[,c(-1,-2,-6)]
```

## Creating Dummy Variables
* Convert all factor variables to dummy variables
```{r message=F}
library(caret)
```

```{r cache=T, message=F}
dummies <- dummyVars(classe ~ ., trainingData)
numData <- as.data.frame(predict(dummies, trainingData))
```

## Zero and Near-Zero Variance Predictors
* Remove all covariates from model consideration which have zero or near-zero variance
```{r cache=T, message=F}
nzv <- nearZeroVar(numData, saveMetrics= TRUE)
dim(numData)

nzv <- nearZeroVar(numData)
filteredData <- numData[, -nzv]
dim(filteredData)
```

## Data Cleansing
* Replaces all missing values (NA) with 0

```{r cache=T, message=F}
for (j in 1:length(names(filteredData))) {
        for (i in 1:nrow(filteredData)) {
              if (is.na(filteredData[i,j])) {filteredData[i,j] <- 0}  
        } 
}
```

## Identifying Correlated Predictors
* Remove highly correlated covariates from model consideration

```{r,  cache=T, results="markup", message=F}
dataCor <-  cor(filteredData)
highCorr <- sum(abs(dataCor[upper.tri(dataCor)]) > .999)
summary(dataCor[upper.tri(dataCor)])

highlyCorData <- findCorrelation(dataCor, cutoff = .75)
filteredData <- filteredData[,-highlyCorData]
dataCor2 <- cor(filteredData[,sapply(filteredData, is.numeric)])
summary(dataCor2[upper.tri(dataCor2)])
```

```{r results='hide'}
numCov <- dim(filteredData)[2]
```
* Before modeling, we are left with `r numCov` covariates for model consideration

## Partition Training and Testing Sets 

```{r, echo=TRUE, cache=T, results="hide",error=FALSE,warning=FALSE, message=F}
classe <- as.vector(trainingData[,"classe"])

trainingData <- cbind(classe, filteredData)

set.seed(1)
inTrain <- createDataPartition(y = trainingData$classe, 
                               p=0.5,
                               list = FALSE)
training <- trainingData[inTrain,]
testing <- trainingData[-inTrain,]
dim(training); dim(testing)
```

# Modeling

## Model Building
```{r, echo=T, cache=T, message=F}
cvCtrl <- trainControl(method = "repeatedcv", repeats = 3)

set.seed(1)
rfTune <- train(classe ~ .,
                data = training, 
                method = "rf",
                ntree=50,
                trControl = cvCtrl
                )
```

## Variable Importance
* Looking at the variables with the highest importance in the Random Forest Model
```{r results='markup'}
varImp(rfTune)
```

* Plotting two of most important variables to the model
```{r}
qplot(x = pitch_forearm,y = num_window,data = training,colour = classe)
```

```{r, echo=T, cache=T, results="markup", message=F, warning=F}
plot(randomForest(classe ~ ., data=training, keep.forest=FALSE, ntree=50), log="y")
```

## Model Testing for Out-of-Sample Error
* Using cross-validation by splitting the training data into a training and testing set. The 50% training set was used to build the Random Forest model. The 50% testing set, which is considered new data to the model, is used to calculate the out-of-sample error.
* The expected out-of-sample error is calculated as the misclassifications made by the model on new data (testing data)

```{r, echo=T, cache=T, results="markup",message=F}
conf <- confusionMatrix(testing$classe, predict(rfTune, testing))
conf$table
etable <- data.frame(actual=testing$classe,
                     predicted=predict(rfTune, testing))
eRate <- nrow(etable[!etable$actual==etable$predicted,]) / nrow(etable)
```
* From this we can expect the out-of-sample error to be `misclassifications / testing cases`, which in this case is `r eRate`.

# Model Scoring New Data

## Pre-Processing New Data
* Format all number based variables as numeric
* Remove variables `X`,`user_name`, and `new_window` from new testing data as they were not used to build the model

```{r}
for (i in 1:length(names(testingData))) {
        if (grepl("dumbbell|forearm|arm|belt",names(testingData)[i])){
                testingData[,i] <- as.numeric(testingData[,i])
        }
}

testingData <- testingData[,c(-1,-2,-6)]
```

## Data Cleansing
* Replaces all missing values (NA) with 0

```{r warning=FALSE, error=FALSE, cache=T, results="hide"}
for (j in 1:length(names(testingData))) {
        for (i in 1:nrow(testingData)) {
              if (is.na(testingData[i,j])) {testingData[i,j] <- 0}  
        } 
}
```

## Creating Dummy Variables
* Convert all factor variables to dummy variables

```{r cache=T, message=F}
dummies <- dummyVars( ~ ., testingData, levelsOnly = TRUE)
numData <- as.data.frame(predict(dummies, testingData))
```

## Predicting with Random Forest Model
* Predicts the `classe` for each of the 20 testing cases
```{r}
answers = as.character(predict(rfTune, numData))

pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(answers)
```


